\texttt{\emph{Incertae sedis}}\texttt{ taxa in supertree construction}

### 2018-03-01 doc/incertae_sedis_latex2.tex
Update.

The OpenTree project seeks to build a comprehensive supertree tree
covering all of life. The approach is to combine information in published
phylogenies with a comprehensive taxonomy that supplies taxon names.
We describe an extension to our supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

@@

In contrast, input phylogenies only have labels on their leaf nodes.
Furthermore, and the set of labels $\mathcal{L}(i)$ on input tree
$T_{i}$ need not include all leaf labels, and in practice always
contains only a relatively small part of $\mathcal{L}$.

@@

@@

\section{Results}
\begin{lyxcode}
\begin{comment}
\begin{lyxcode}
?ott6145835~{[}6145835{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}

?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~

?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}

?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

?ott564110~{[}564110{]}~<-~ott939130~{[}939130{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~

?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~

?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~

?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}

?ott441413~{[}441413{]}~<-~ott1014022~{[}1014022{]}~placed~under~ott571842~{[}571842{]}~<-~ott997381~{[}997381{]}~<-~ott1014022~{[}1014022{]}~

?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~

?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~

?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~
\end{lyxcode}
\end{comment}
\end{lyxcode}
Should we do this? We could say: 
\begin{itemize}
\item we placed 14 incertae sedis taxa.
\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.

>>>>

The OpenTree project seeks to build a comprehensive supertree covering
all of life. The approach is to combine information in published phylogenies
with a comprehensive taxonomy that supplies taxon names \citep{redelings2017supertree}.
We describe an extension to our supertree method so that we can resolve
taxonomic uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa. We describe an updated semantics for conflict between
phylogeny edges and taxonomy edges and an updated semantics for assigning
taxonomy names to nodes in the synthetic tree.

@@

In contrast, input phylogenies only have labels on their leaf nodes.
Furthermore, the set of labels $\mathcal{L}(i)$ on input tree $T_{i}$
need not include all leaf labels, and in practice always contains
only a relatively small part of $\mathcal{L}$.

@@

\begin{comment}
While we get Regulus back, we do not end up placing it in Sylviidae
because Sylviidae is broken. It conflicts with about 20 different
trees.
\end{comment}

@@

\section{Results}

\subsection{Case 1}

taxonomy = (((a1,a2)A,(b1,b2)B)AB,(c1,c2)?C,D)root;

\subsubsection{If we place $c$ within $A$}

tree1 = (((a1,c1),b1),d1);

synth-with-is: ((((a1,(c1,c2)C),a2)A,(b1,b2)B)AB),(d1)D)

placement: C within A <- AB

synth-no-is: ((a1,(c1,c2)C),a1,(b1,b2)B,(d1)D)

\subsubsection{If we place $c$ within $A$ but break $C$}

tree1 = ((a1,c1),(a2,c1),b1)

synth-with-is: ((((a1,c1),(a2,c2))A,(b1,b2)B)AB, (d1)D)

synth-no-is: ??

placement: 
\begin{itemize}
\item c1 <- (broken) C within A <- AB.
\item c2<- (broken) C within A <- AB
\end{itemize}
result: we lose $C$, but keep $A$ and $AB$.

\subsection{Case 2: nested is}

taxonomy = (((a1,a2)A,(b1,b2)B),( ( (c11,c12)C1,(c21,c22)C2,(ci1,ci2)?CI
)?C) ,(d1)D)

tree1 = ((a1,(c11,(ci1,c12))),b1)

synth-with-is: places CI wihin C1, and C within A.

\subsection{Case 3: interleaving taxa}

taxonomy = ((a1,a2)A,(b1,b2)?B,(c1,c2)?c)

input tree: ((b1,c1),(b2,c2))

synth-with-is: ((a1,a2)A,((b1,c1),(b2,c2))B) or ((a1,a2)A,((b1,c1),(b2,c2))C)

Result: we name the interleaved taxon $B$ or $C$, but cannot name
it both names.

\subsection{Case 4: creation of monotypic taxa}

see test cases.
\begin{lyxcode}
\begin{comment}
\begin{lyxcode}
\textbf{Whale:}~?ott6145835~{[}6145835~Brandtocetus/genus{]}~<-~ott698424~{[}698424~Cetacea/order{]}~placed~under~ott4942423~{[}4942423~Cetotherium~/~genus{]}~<-~ott698424~{[}698424~Cetacea/order{]}

\textbf{Whale:}~?ott6145873~{[}6145873{]}~<-~ott698424~{[}698424~Cetacea~/~order{]}~placed~under~ott4942423~{[}4942423{]}~<-~ott698424~{[}698424{]}~

-~Cetacea~broken.

\textbf{Insect:}~?ott18090~{[}18090{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}

\textbf{Insect:}~?ott190269~{[}190269{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

\textbf{Insect:}~?ott349638~{[}349638{]}~<-~ott672501~{[}672501{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

\textbf{Insect:}~?ott1009313~{[}1009313{]}~<-~ott672501~{[}672501~Reduviidae~/~family{]}~placed~under~ott275501~{[}275501{]}~<-~ott672501~{[}672501{]}~

-~Reduviidae~broken.

\textbf{Insect:~}?ott564110~{[}564110{]}~<-~ott939130~{[}939130~Thyrididae~/~family{]}~placed~under~ott757344~{[}757344{]}~<-~ott939130~{[}939130{]}~

\textbf{Plant:}~?ott679485~{[}679485{]}~<-~ott1074933~{[}1074933~Lamioidea~/~subfamily{]}~placed~under~ott534796~{[}534796{]}~<-~

\textbf{Plant:}~?ott729184~{[}729184{]}~<-~ott1074933~{[}1074933~Lamioidea~/~subfamily~{]}~placed~under~ott534796~{[}534796{]}~<-~ott1074933~{[}1074933{]}~

ott1074933~{[}1074933{]}~

-~Lamioidea~\emph{not}~broken

\textbf{Plant:}~?ott5144555~{[}5144555{]}~<-~ott23373~{[}23373{]}~placed~under~ott447802~{[}447802{]}~<-~ott23373~{[}23373{]}

-~Orobanchaceae~broken

\textbf{Fly:}~?ott441413~{[}441413~Rhodesiella~lungliensis/species{]}~<-~ott1014022~{[}1014022~Rubiaceae{]}~placed~under~ott571842~{[}571842~Pavetteae~/~tribe{]}~<-~ott997381~{[}997381~Ixoroidea/subfamily{]}~<-~ott1014022~{[}1014022~Rubiaceae/family{]}

\textbf{Plant:}~?ott31033~{[}31033{]}~<-~ott627035~{[}627035{]}~placed~under~ott6019314~{[}6019314{]}~<-~ott627035~{[}627035{]}~

\textbf{Bacteria:}~?ott4018918~{[}4018918{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~

\textbf{Bacteria:}?ott4792759~{[}4792759{]}~<-~ott5248084~{[}5248084{]}~placed~under~ott167767~{[}167767{]}~<-~ott5248084~{[}5248084{]}~

-~Erysipelotrichaceae~broken
\end{lyxcode}
\end{comment}
\end{lyxcode}
Should we do this? We could say: 
\begin{itemize}
\item there are 331 incertae sedis taxa that are mentioned in input trees.
\begin{itemize}
\item what is the kingom / phylum / class / order / genus / species
\end{itemize}
\item we placed 14 incertae sedis taxa inside a sister taxon.
\item we confirmed $n$ incertae sedis taxa as being separate from all sister
(sampled) taxa??
\item we avoiding breaking 8 taxa that had IS taxa placed inside them.
\item we allowed $z_{1}$ new taxa into the synthesis tree that were incertae
sedis.
\item we allowed $z_{2}$ new taxa into the synthesis tree that are marked
as extinct.
\item some nodes have as many as $w$\emph{ incertae sedis} children, making
them unbrowseable when incertae sedis children are not excluded.
\item $v_{1}$ input trees were previously excluded \emph{entirely} because
they are nested within in an incertae sedis taxon.
\item $v_{2}$ input trees were previously excluded \emph{partially} because
they are nested within in an incertae sedis taxon.
\end{itemize}
Currently the numbers $z_{1}$ and $z_{2}$.
